---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
<br><br>

```{r setglobaloptions, echo=FALSE}
# set up the environment; force knitr install in case it is not installed on 
# reviewer's machine

library(knitr)
opts_chunk$set(echo = TRUE, message = FALSE)
options(scipen = 1, digits = 4)
```

## Loading and preprocessing the data   

After downloading and saving the zip file in the current working directory, we
extract the activity monitoring data stored in the activity.csv file. We then 
load the dataset into a tbl_df data frame. Finally, we convert the 'date' 
column from factor type to date type.

```{r loaddata}
library(lubridate)
library(dplyr)

file <- 'activity.csv'
tdf <- tbl_df(read.csv(file))
tdf$date <- ymd(tdf$date)  # convert factor to POSIXct date type
output some basic dataset summary data just in case something happens to the 
```
<br><br>

We output some summary data just in case something happens to the CSV file so 
the person reviewing this report can quickly assess whether or not the data
source is the same before even starting to look into all the details of the 
analysis.

```{r summary_data}
str(tdf)
summary(tdf)
```
<br><br><br>

## What is the mean total number of steps taken per day?   

We create a histogram of the total number of steps taken each day after 
discarding any row with a 'NA' value in the 'steps' column.   

```{r histogram}
tsd <- tdf %>% group_by(date) %>% summarize(total_steps_day = sum(steps, na.rm = TRUE))
hist(tsd$total_steps_day, main = 'Histogram of total steps each day', xlab = 'Total steps')
```
<br><br>

The code used to produce the mean and median is:

```{r mean_median}
# note that NAs have already been removed by the code above in the call to sum()
mmd <- tsd %>% summarize(mean = mean(total_steps_day), median = median(total_steps_day))
```

The **mean** total number of steps taken per day is **`r mmd$mean`** and the 
**median** total number of steps taken per day is **`r mmd$media`**.  
<br><br><br>

## What is the average daily activity pattern?

The plot showing the average number of steps for each interval in a day across 
all days is produced by the following code:   

```{r timeseries_plot}
ctdf <- complete.cases(tdf)
iads <- tdf[ctdf, ] %>% group_by(interval) %>% summarize(davg = mean(steps))
plot(iads$interval, iads$davg, type = 'l', main = 'Avg Daily Activity', xlab = '5-min interval', ylab = 'Avg Steps')
```


```{r max_interval}
max_interval <- iads[max(iads$davg) == iads$davg, 'interval']
```
The 5-minute interval that contains the maximum number of steps on average across all the days in the dataset is **`r max_interval`**.
<br><br><br>

## Imputing missing values

The following code displays the total number of rows in the dataset that contain 
at least 1 missing value.

```{r totalNA}
totalNAs <- sum(is.na(tdf))
```

The total number of rows missing some values thus is **`r totalNAs`**.

The strategy we will use to fill in all the missing values consists in taking
the average number of steps for a particular interval over the whole period under
consideration and use that value in lieu of the missing one. These averages are
already available as they have been computed above and stored in the 'iads' 
variable. For convenience, we repeat the code here:

```{r interval_avg}
iads <- tdf[ctdf, ] %>% group_by(interval) %>% summarize(davg = mean(steps))
```

We now create a new dataset equal to the original one but with the missing 
data filled in. To do so efficiently, we need the data.table library.

```{r missing_data_replace}
# use data.table for in place mutation as it's more elegant than a bunch of
# loops and avoids more awkward and convoluted syntax (the S people didn't
# know how to design a programming language properly; what a shame)
library(data.table)

dt <- as.data.table(tdf)
dt$steps <- as.numeric(dt$steps)
dt[is.na(steps) & interval %in% iads$interval, steps := iads$davg]
```

We now make a new histogram of the total number of steps taken each day.

```{r new_histogram}
# since NAs have been removed, for efficiency no need to use na.rm arg
ntsd <- dt %>% group_by(date) %>% summarize(total_steps_day = sum(steps)) 
hist(ntsd$total_steps_day, main = 'Histogram of total steps each day (no NAs)', xlab = 'Total steps')
```
<br><br>

Finally, we compute the mean and median again but this time based on the new 
values used to replace NAs:

```{r new_mean_median}
nmmd <- ntsd %>% summarize(mean = mean(total_steps_day), median = median(total_steps_day))
```

The new **mean** total number of steps taken per day is **`r nmmd$mean`** and 
the new **median** total number of steps taken per day is **`r nmmd$media`**.  Interestingly, 
these values are quite different from the previous estimates. It appears that 
the impact of imputing missing data affects the skew of the distribution.
<br><br><br>


## Are there differences in activity patterns between weekdays and weekends?

```{r wkdays_wkends}
library(ggplot2)
dt$daytype <- as.factor(ifelse(weekdays(dt$date) %in% c("Saturday", "Sunday"),'weekend','weekday'))
iads_by_daytype <- dt %>% group_by(interval, daytype) %>% summarize(davg = mean(steps))
qplot(interval, davg, data = iads_by_daytype, facets = daytype.~, geom = 'path', main = 'Weekday vs weekend activity patterns') 
```